Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-07-12 manuscript/manuscript.qmd
all sections of the discussion is ok

Typically, sleep detection methods are applied in two contexts: either
to night recordings or to 24-hour recordings. In night recordings, it is
possible to derive sleep quality parameters like SE and LPS because the
SPT is already known because it is inferred from the length of the
recording[@conley2019; @patterson_40_2023]. On the other hand, when
sleep detection methods are applied to 24-hour recordings, most methods
do not have the ability to infer the SPT with sleep
diaries[@girschik2012]. Consequently, these methods are unable to
generate certain sleep quality parameters that rely on the
SPT[@doherty2017; @anderson2014]. We have attempted to circumvent this
issue by employing models able to distinguish between in-bed awake time
and in-bed asleep time over a 24h recording making our models capable of
estimating all commonly used sleep quality parameters. Van Hees et
al.[@van_hees_estimating_2018] have developed an algorithm to identify
SPT from wrist-worn devices which was recently validated by Plekhanova
and colleagues[@plekhanova2023]. which in combination with other
algorithms or models on said SPT can then be used infer further sleep
quality parameters.

Strengths and limitations It is not always clear the methods for
detecting sleep from accelerometer data is

It is unclear whether Johansson et al.[@johansson_development_2023]
provide a method for detecting the in-bed time which hinders its
applicability for researchers that want to ...

external validation dataset/generalization

expand the feature set (Palotti)

In conclusion, our study has advanced the methods for estimating sleep
from thigh-worn accelerometers by evaluating various models for
predicting in-bed and sleep time. Our results indicate that all
sequential models and the multiclass model performed well at predicting
in-bed time. However, all models showed difficulty in correctly
classifying awake epochs during time in bed, highlighting the challenge
of automating the detection of in-bed awake periods. The application of
median filters improved the performance metrics of all models,
suggesting that this approach may be beneficial in future studies.
Despite these promising results, our findings also underscore the need
for caution when applying these models to individual-level sleep
assessments due to the high variability in derived sleep quality
parameters. Future research should focus on refining these models to
enhance their precision and reliability for individual sleep
assessments.

Our open-source algorithm and the inferred sleep parameters will open
the door to future studies on sleep and sleep architecture using
large-scale accelerometer databases.

>>>>

Typically, sleep detection methods are applied in two contexts: either
to night recordings or to 24-hour recordings. In night recordings, it is
possible to derive sleep quality parameters like SE and LPS because the
SPT is already known because it is inferred from the length of the
recording[@conley2019; @patterson_40_2023]. On the other hand, when
sleep detection methods are applied to 24-hour recordings, most methods
do not have the ability to infer the SPT with sleep
diaries[@girschik2012]. Consequently, these methods are unable to
generate certain sleep quality parameters that rely on the
SPT[@doherty2017; @anderson2014]. To overcome this limitation, we have
incorporated models that can differentiate between in-bed awake time and
in-bed asleep from out-bed awake time over a 24-hour recording. This
approach allows our models to estimate all commonly used sleep quality
parameters. Van Hees et al. [@van_hees_estimating_2018] have proposed an
algorithm to determine SPT from data collected by wrist-worn devices.
This algorithm was recently validated by Plekhanova and her team
[@plekhanova2023]. By combining this algorithm with other methods,
further sleep quality parameters can be inferred based on the identified
SPT. Van Hees et al.[@van_hees_estimating_2018] reported good agreements
and low mean differences compared to self-report and PSG on SPT,
findings later confirmed by Plekhanova and colleagues. However, they
also observed poor agreement with LPS and Wake After Sleep Onset (WASO).
They found low reliability with PSG, indicating difficulties in
detecting wakefulness during in-bed time. These challenges parallel
those we experienced in our study.

The study boasts several strengths, including the capacity to derive
in-bed awake and asleep times from out-of-bed data, thereby allowing for
the extraction of vital sleep quality parameters. Furthermore, the
research benefits from evaluating multiple nights per subject, providing
valuable information into intra-subject sleep variability. However,
certain limitations exist. The use of ZM, which isn't recognized as a
gold standard, could potentially compromise our findings' validity.
Future research could consider using PSG as a reference for methods
similar to ours, despite its limitations, for a more accurate
comparison. Moreover, our models weren't validated using an external
dataset, a process that would have showcased their broader
applicability. Hence, our conclusions remain confined primarily to
children.

In conclusion, our study contributes to the ongoing efforts to improve
sleep estimation methods using thigh-worn accelerometers. We evaluated
different machine learning models and configurations for predicting
in-bed and sleep times and their corresponding sleep quality parameters.
While the sequential models generally demonstrated excellent performance
in predicting in-bed time, they faced challenges in accurately
distinguishing between sleep and wake epochs during in-bed time.
Nevertheless, their performance could be substantially improved through
the application of median filters. Among all models and configurations
evaluated, the XGBoost model exhibited superior performance across all
performance metrics, including epoch-to-epoch predictions and sleep
quality parameter derivatives. Our research also highlighted the current
limitations of sleep detection methods, such as challenges in
effectively detecting wake periods during in-bed time and the need for
further improvements to increase the precision of individual sleep
assessments. We believe our work lays the groundwork for future research
to further refine and improve the performance of these models,
contributing to a more precise and accurate evaluation of sleep patterns
and quality using thigh-worn accelerometers.
